The role of advection versus diffusion in intraventricular transport

### 2025-02-13 paper/main.tex
update figure captions, clean mesh msh-to-xdmf code

The morphology and geometry of the brain ventricles vary across zebrafish
individuals, and under physiological and pathological conditions. 
Motivated by an interest in how ventricular geometry impacts solute distribution,
we also consider four variations in the geometry \fixme{(percentage reduction in cross-sectional areas):
by shrinking the fore-mid brain connection (66 \%), 
the middle ventricle (33 \%), the mid-hind brain connection (43 \%), and all three simultaneously.} 
We used Blender~\cite{Community2018BlenderPackage} to modify the brain ventricles geometry.
% Cross-sectional areas of ducts normal to flow direction
% Modified region       Original mesh        Deformed mesh
% Fore                  3.36e-3 mm^2         1.15e-3 mm^2
% Mid                   6.11e-3 mm^2         4.08e-3 mm^2
% Hind                  2.91e-3 mm^2         1.67e-3 mm^2

@@

\subsection*{Solution strategy and software implementation}
Owing to the one-way coupling between the velocity field $\uu$
and the concentration $c$, the governing equations for the
CSF flow and the concentration can be solved sequentially when modeling solute transport.
\fixme{For the cilia-only/no-cardiac flow scenario, the velocity quickly reaches steady-state.}
In this case, we first solve the \fixme{steady-state} Stokes equations~\eqref{eq:stokes_eqs}
for $\uu$ and $p$, and then use the velocity field $\uu$ as input when solving
the advection-diffusion equation~\eqref{eq:adv_diff_strong}.
For the other flow scenarios, the Stokes equations are solved for one
period of the pulsatile cardiac motion. The solution of this one period
is then used periodically as input for the advection-diffusion equation.

The numerical solution schemes for the Stokes equations and the
advection-diffusion equation were implemented and solved numerically with DOLFINx,
the Python interface of the finite element software
FEniCSx~\cite{TheFEniCSProject2024FEniCSxDocumentation}, with
petsc4py~\cite{Dalcin2011ParallelPython} as linear algebra backend.
A direct solver using MUMPS~\cite{Amestoy2011Mumps} was used to solve the
linear systems resulting from the discretized Stokes equations.
For the discretized advection-diffusion equation, we used an iterative
linear solver based on GMRES~\cite{Saad1986GMRES:Systems} with block Jacobi
preconditioning~\cite{Jacobi1845UeberGleichungen}.
Numerical experiments verified the accuracy of the model implementation
(\Cref{sec:appB_verification}). The simulation code is openly
available~\cite{zenodo-link-to-all-code}.

@@

\subsection*{Photoconversion data analysis}
All data analysis was performed in MATLAB and the codes are available in
the repository of this article: \url{repo-link}.

@@

Next we consider the baseline model, simulating flow driven by both cilia
and cardiac pulsations. We observe persistent flow features,
with vortex-structures in the anterior and middle ventricles
and directional flow in the posterior ventricle (\Cref{fig:fig1}h--j,
\fixme{Supplementary video 1}). These characteristic patterns suggest
that the flow in the anterior and middle ventricles is majorly driven
by cilia, while the cardiac pressure pulsations dominate
within the posterior ventricle and in the ducts connecting the ventricular
compartments. Over one cardiac cycle, the CSF flow
speed reaches up to 27.7 \textmu m/s, dorsally in the posterior part
of the middle ventricle. This peak is attained at the time instant
coinciding with the time of peak caudal cardiac pulsatile flow, since 
both the tangential traction induced by the cilia and the pulsatile
pressure-driven flow are aligned with the rostrocaudal 
axis. Using the observed maximum velocity magnitude and choosing the height of
the middle ventricle ($110$ \textmu m) as a characteristic length
scale yields a Reynolds number $\mathrm{Re} \approx 0.004$,
justifying the assumption of Stokes flow. In summary, the computational
CSF flow model reproduces the flow features observed \emph{in vivo}~\cite{Olstad2019CiliaryDevelopment}.

\subsection*{Ventricular solute transport is advection-dominated for larger molecules}
We next explore the impact of these CSF flow patterns and magnitudes
on solute transport within the brain ventricles.
To this end, we simulate the distribution and evolution
of different solutes subjected to photoconversion within a region of
interest in the dorsal region of the middle ventricle (ROI 1)
(\Cref{fig:fig2}a). We consider the spread of
a solute concentration with a given molecular diffusion coefficient,
corresponding to exosomes ($D_1 = 1.63 \times 10^{-12}$ m$^2$/s),
a circa 90 kDa protein STM-GFP ($D_2 = 5.75 \times 10^{-11}$ m$^2$/s),
and the 26 kDa photoconvertible protein  Dendra2 ($D_3 = 1.15 \times 10^{-10}$ m$^2$/s).
As expected, transport occurs more
rapidly with increasing diffusion coefficient. Interestingly, the
distribution patterns differ between the solutes.
For the smallest diffusion coefficient, representing
exosomes, the advective transport by the CSF flow dominates diffusion
(\Cref{fig:fig2}b). The vortex structures of the velocity field
carry over into the concentration field, indicating that transport
along the streamlines is more rapid than the diffusion across the
streamlines. The balance between diffusion and advection shifts with
increasing diffusion coefficients: the CSF flow structures become more
obfuscated, and the solute spreads more uniformly throughout the
ventricular geometry (\Cref{fig:fig2}c, d). Notably, the
smaller solutes with higher diffusion coefficients spread more quickly
to regions distal to the photoconversion site.
\begin{figure}%[H]
    \centering
    \includegraphics[width=0.95\textwidth]{graphics/figure2_simulation_results_different_coeffs.png}
    \caption{Simulated distribution and evolution after photoconversion of
    exosomes ($D_1$), STM+GFP ($D_2$) and Dendra2 ($D_3$).
    \textbf{a.} Geometry-centered clip of the ventricles mesh showing the locations of the ROIs.
    \textbf{b--d.} Slice of the concentration field simulated
    with diffusion coefficients $D_1$, $D_2$ and $D_3$ for eight time instants covering the
    time range from 45 seconds to 790 seconds, corresponding to 100 to 1750 cardiac cycles.
    Slices are taken in the $xz$-plane ($y=0.140$ mm, the center of the geometry).
    The colorbar is scaled so the maximum is the maximum concentration of the Dendra2 
    concentration field slice at the last time instant.
    \textbf{e--g.} Mean concentration $\cbar$ in each region of interest (ROI)
    as function of time for diffusion coefficient $D_1$, $D_2$ and $D_3$, respectively. 
    \textbf{h.} The time $\hat{t}$ before the mean concentration $\cbar$ exceeds a
    threshold value of $0.25$ (ROIs 2-4) or $0.10$ (ROIs 5, 6) in each simulation setup.
    Note that there is no bar for $D_1$ in ROI 6, because the threshold value was never exceeded.}
    \label{fig:fig2}
\end{figure}

To quantify the transport of solute, we plot the profile of the mean
concentrations $\cbar_i(t)$ as functions of time
(\Cref{fig:fig2}e--g) and
the times $\hat{t}$ when the mean concentrations $\cbar_i$ first exceed
a threshold value $\hat{c}$ (\Cref{fig:fig2}h)
in 6 ROIs (\Cref{fig:fig2}a).
The dynamics at the site of photoconversion (ROI 1)
is logarithmic and follows from imposing \eqref{eq:photoconversion_curve}.
For ROIs 2--6, we observe faster kinetics with higher diffusion coefficients
(\Cref{fig:fig2}e--g),
which results in smaller times-to-threshold
(\Cref{fig:fig2}h).
For all three diffusion coefficients, the dynamics of ROI 5
is significantly delayed compared to ROI 2--4,
and ROI 6 is the slowest. Notably, for $D_1$ the mean concentration
in ROI 6 is close to zero at the end of simulation.
To quantify the balance of advection versus diffusion as transport mechanisms,
we calculate the Péclet numbers $\mathrm{Pe}_1=519$,
$\mathrm{Pe}_2=14.7$ and $\mathrm{Pe}_3=7.4$ for $D_1$, $D_2$ and $D_3$, respectively.
Since $\mathrm{Pe}>1$, this indicates that advection is the
dominant transport mechanism on a \fixme{global} scale.

@@

We obtained very similar kinetics curves (\Cref{fig:fig3}a, b)
and time-to-threshold values (\Cref{fig:fig3}c, d) when comparing our
experiments (\Cref{fig:fig3}a, c) with
simulation (\Cref{fig:fig3}b, d).
Importantly, since the initial condition of the simulation is a mean concentration of zero,
the mean concentration $\cbar$ can be interpreted as the change in mean concentration and
is thus a similar metric to $\Delta F$.
The final values of $\cbar$ in our simulations 0.76, 0.72, 0.82, 0.56 and 0.35 (ROIs 2--6, respectively)
were generally higher than the experimental values 0.63, 0.71, 0.67, 0.29 and 0.18 for $\Delta F$.
Notwithstanding similarities, we observed small differences between the experiments and simulation.
For instance, transport dynamics in ROIs 2--6 were faster in the simulation and reached higher final values.
Moreover, the $\cbar$ profiles of ROIs 3--4 were opposite order in
the simulations as compared to the experiments.
Despite observing small differences, the \emph{in vivo} experiments and
simulations were similar enough to convincingly validate the model.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure3_compare_experiments_and_simulations.png}
    \caption{
    \textbf{a.} Relative change in fluorescence intensity $\Delta F$ over time
    in each region of interest (ROI)
    observed during experiments of photoconversion of Dendra2 proteins in ROI 1.
    \textbf{b.} Mean concentration $\cbar$ as function of time in each ROI in a
    simulation of Dendra2 protein transport with
    the diffusion coefficient $D_3$.
    \textbf{c.} Time to $\Delta F$ reaches a threshold value of 0.25 in
    one representative zebrafish larva.
    Time to $\cbar$ reaches a threshold value of 0.25 for
    a transport simulation with diffusion coefficient $D_3$.
    We have excluded the posterior-most regions of the geometries,
    because the values of $\Delta F$ and $\cbar$ never exceeded the threshold
    value of 0.25 in these regions.}
    \label{fig:fig3}
\end{figure}
% Final Delta F values
%0.6322650156762044 ROI 2
%0.7059449124175479 ROI 3
%0.6694500367160006 ROI 4
%0.28986700094045437 ROI 5
%0.18155055624229066 ROI 6
% Final mean c values ROIs 2-6: 0.7637806  0.72025859 0.81903216 0.56204164 0.34926411

\subsection*{Absence of ciliary motion affects local protein distribution}
Since CSF flow is highly disturbed in the absence of motile cilia, we
want to assess the impact of cilia motion on solute distribution. To this end,
we measured the movement of photoconverted Dendra2 signal
in control and motile-cilia mutant zebrafish (\emph{smh}).
By plotting the means of $\Delta F$ (\Cref{fig:fig4}a)
and the time to threshold
across the control and mutant cohorts (\Cref{fig:fig4}b),
we identified a significant
difference in Dendra2 distribution in ROIs 2 and 4, but not in ROIs 3, 5, and 6. In ROIs 2 and 4,
which measure Dendra2 distribution downstream of the photoconversion site,
we observed a slower increase of fluorescence intensity,
resulting in $\sim 45\%$ (ROI 2) and $\sim 25\%$ (ROI 4)
higher mean times-to-threshold in the \emph{smh}
mutants compared to controls (\Cref{fig:fig4}b).

Next, to identify whether our model would recapitulate these findings,
we simulated protein transport by using the only-cardiac/no-cilia flow model 
and compared it with the baseline model
(both cilia and cardiac motion) (\Cref{fig:fig4}c).
Similar to our experiments, we observed slower dynamics in simulations for ROIs 2 and 4,
but small changes for ROIs 3 and 6.
Interestingly, the difference was even greater for the simulations,
with the time-to-threshold in ROI 2 being 90\% higher in absence of cilia-mediated flow.
Moreover, ROI 5 exhibits very different $\cbar$ profiles for the two
configurations in our computational model,
but not in our experimental data.
Altogether, our findings identify that diffusion plays an important
role for solute movement to regions upstream in the cilia flow fields or to distant ventricles.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $F$ over
    time in each region of interest (ROI) for control (black lines) and \emph{smh} mutant (red lines) fish. 
    \textbf{b.} Times when $\Delta F$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for control (gray dots) and
    \emph{smh} mutant (red dots) fish.
    The thick black lines denote mean values of the datasets, and $p$-values
    were calculated using a Mann-Whitney U test with a confidence level of 95\%.
    \textbf{c.} Mean concentration $\cbar$ as function of time in each ROI,
    simulated using the baseline model (black lines) and the cardiac-only/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in both panels \textbf{a} and \textbf{c}, the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

\subsection*{Loss of cilia motility in specific subpopulation slows down transport}
We previously identified three distinct ciliated cell lineages in the
zebrafish larval brain~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain}.
To identify the role of these lineages on CSF transport, we simulate transport
upon region-specific cilia paralysis by modifying the boundary conditions
of the CSF flow model. Three separate cilia-covered regions were considered separately:
the telencephalic (anterior) ventricle, the dorsal part and the ventral part
of the diencephalic (middle) ventricle (\Cref{fig:fig5}a).
The cilia tangential traction boundary condition is for each case
changed to a free-slip boundary condition in paralyzed region,
prescribing zero tangential forces to
represent absence of cilia motion.
We simulate CSF flow modeled by the Stokes equations with the
modified boundary conditions and use the resulting
velocity fields in transport simulations.

We identified that the velocities were most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity dropped from 27.7 \textmu m/s to 8.0 \textmu m/s.
\emph{In vivo}, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic had minimal
impact and resulted in maximum velocities of 27.7 \textmu m/s or 27.2 \textmu m/s, respectively.
Changes in the transport dynamics of Dendra2 were most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit changes in the dynamics of $\cbar$ (\Cref{fig:fig5}b).
Compared to the baseline model results, we observed a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
On the other hand, removal of telencephalic cilia did not alter the evolution of the
mean concentration $\cbar$ with time (\Cref{fig:fig5}b).
Finally, without ventral cilia in the diencephalic ventricle,
more solute has spread towards the anterior ventricle (ROI 5) and
less solute has spread towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions where loss of cilia motility is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    loss-of-cilia-motility scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

\subsection*{Modification of ventricles geometry impacts protein distribution}
Variations in ventricular morphology impact CSF flow patterns.
To assess how ventricular morphologies influence solute distribution,
we used four alternative brain ventricles meshes:
one where the connection between the anterior and middle ventricles is constricted;
one where the connection between the middle and posterior ventricles is constricted;
one where the middle ventricle is shrunk in the lateral direction;
and the combination of the above three modifications (\Cref{fig:fig6}a).
We simulate CSF flow in the modified geometries with the baseline flow model
and use the resulting velocity field to simulate transport of Dendra2
(\Cref{fig:fig6}b).

Constricting the connection between the anterior and middle ventricle leads to
slower transport to the anterior ventricle (ROI 5),
and a slightly higher mean concentration in ROIs 2--4 and 6, compared to the original geometry.
On the other hand, constricting the connection between the middle and the posterior ventricle
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a shrunk middle ventricle.
Additionally, for this configuration the transport upstream the flow fields towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.
For the mesh where all three ventricles are modified,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while both the anterior and the posterior ventricles (ROIs 5 and 6) have lower concentration.
Altogether our results identify that the geometry and width of the ventricular ducts matter
for solute transport and may explain why we observed larger differences
in the cilia-deficient model than in the experimental data in
\Cref{fig:fig4}, which have different duct diameters.

\lyng{HH: Quantify differences in terms of percentage changes in the
final mean concentration values, maybe with a bar plot?}
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{Caption. \mk{MK: How do the different geoemtries compare in terms of volume,
    areas where cilia are active?} \lyng{HH: Good question, will add this information.}}
    \label{fig:fig6}
\end{figure}

>>>>

The morphology and geometry of the brain ventricles vary across zebrafish
individuals, and under physiological and pathological conditions. 
Motivated by an interest in how ventricular geometry impacts solute distribution,
we also consider four variations in the geometry \fixme{(cross-sectional area
reduction in parentheses):
by shrinking the fore-mid brain connection (66 \%), 
the middle ventricle (33 \%), the mid-hind brain connection (43 \%), and all three simultaneously.} 
We used Blender~\cite{Community2018BlenderPackage} to modify the brain ventricles geometry.
% Cross-sectional areas of ducts normal to flow direction
% Modified region       Original mesh        Deformed mesh
% Fore                  3.36e-3 mm^2         1.15e-3 mm^2
% Mid                   6.11e-3 mm^2         4.08e-3 mm^2
% Hind                  2.91e-3 mm^2         1.67e-3 mm^2

@@

\subsection*{Solution strategy and software implementation}
Owing to the one-way coupling between the velocity field $\uu$
and the concentration $c$, the governing equations for the
CSF flow and the concentration can be solved sequentially when modeling solute transport.
For the cilia-only/no-cardiac flow scenario, the velocity quickly reaches steady-state since
the viscous forces dominate the inertial forces.
In this case, we first solve the steady-state Stokes equations~\eqref{eq:stokes_eqs}
for $\uu$ and $p$, and then use the velocity field $\uu$ as input when solving
the advection-diffusion equation~\eqref{eq:adv_diff_strong}.
For the other flow scenarios, the Stokes equations are solved for one
period of the pulsatile cardiac motion. The solution of this one period
is then used periodically as input for the advection-diffusion equation.

The numerical solution schemes for the Stokes equations and the
advection-diffusion equation were implemented and solved numerically with DOLFINx,
the Python interface of the finite element software
FEniCSx~\cite{TheFEniCSProject2024FEniCSxDocumentation}, with
petsc4py~\cite{Dalcin2011ParallelPython} as linear algebra backend.
A direct solver using MUMPS~\cite{Amestoy2011Mumps} was used to solve the
linear systems resulting from the discretized Stokes equations.
For the discretized advection-diffusion equation, we used an iterative
linear solver based on GMRES~\cite{Saad1986GMRES:Systems} with block Jacobi
preconditioning~\cite{Jacobi1845UeberGleichungen}.
Numerical experiments verified the accuracy of the model implementation
(\Cref{sec:appB_verification}). The simulation code is openly
available~\cite{zenodo-link-to-all-code}\fixme{Add repo link.}.

@@

\subsection*{Photoconversion data analysis}
All data analysis was performed in MATLAB and the codes are available in
the repository of this article~\cite{zenodo-link-to-all-code}\fixme{Add repo link.}.

@@

Finally, we consider the baseline model, simulating flow driven by both cilia
and cardiac pulsations. We observe persistent flow features,
with vortex-structures in the anterior and middle ventricles
and directional flow in the posterior ventricle (\Cref{fig:fig1}h--j,
\fixme{Supplementary video 1}). These characteristic patterns suggest
that the flow in the anterior and middle ventricles is majorly driven
by cilia, while the cardiac pressure pulsations dominate
within the posterior ventricle and in the ducts connecting the ventricular
compartments. Over one cardiac cycle, the CSF flow
speed reaches up to 27.7 \textmu m/s, dorsally in the posterior part
of the middle ventricle. This peak is attained at the time instant
coinciding with the time of peak caudal cardiac pulsatile flow, since 
both the tangential traction induced by the cilia and the pulsatile
pressure-driven flow are aligned with the rostrocaudal 
axis. Using the observed maximum velocity magnitude and choosing the height of
the middle ventricle ($110$ \textmu m) as a characteristic length
scale yields a Reynolds number $\mathrm{Re} \approx 0.004$,
justifying the assumption of Stokes flow. In summary, the computational
CSF flow model reproduces the flow features observed
\emph{in vivo}~\cite{Olstad2019CiliaryDevelopment}.

\subsection*{Ventricular solute transport is advection-dominated for larger molecules}
We next explore the impact of these CSF flow patterns and magnitudes
on solute transport within the brain ventricles.
To this end, we simulate the distribution and evolution
of different solutes subjected to photoconversion within a region of
interest in the dorsal region of the middle ventricle (ROI 1)
(\Cref{fig:fig2}a). We consider the spread of
a solute concentration with a given molecular diffusion coefficient,
corresponding to exosomes ($D_1 = 1.63 \times 10^{-12}$ m$^2$/s),
a circa 90 kDa protein STM-GFP ($D_2 = 5.75 \times 10^{-11}$ m$^2$/s),
and the 26 kDa photoconvertible protein  Dendra2 ($D_3 = 1.15 \times 10^{-10}$ m$^2$/s).
As expected, transport occurs more
rapidly with increasing diffusion coefficient. Interestingly, the
distribution patterns differ between the solutes.
For the smallest diffusion coefficient, representing
exosomes, the advective transport by the CSF flow dominates diffusion
(\Cref{fig:fig2}b). The vortex structures of the velocity field
carry over into the concentration field, indicating that transport
along the streamlines is more rapid than the diffusion across the
streamlines. The balance between diffusion and advection shifts with
increasing diffusion coefficients: the CSF flow structures become more
obfuscated, and the solute spreads more uniformly throughout the
ventricular geometry (\Cref{fig:fig2}c, d). Notably, the
smaller solutes with higher diffusion coefficients spread more quickly
to regions distal to the photoconversion site.
\begin{figure}%[H]
    \centering
    \includegraphics[width=0.95\textwidth]{graphics/figure2_simulation_results_different_coeffs.png}
    \caption{Simulated distribution and evolution after photoconversion of
    exosomes ($D_1$), STM+GFP ($D_2$) and Dendra2 ($D_3$). \\
    \textbf{a.} Geometry-centered clip of the ventricles mesh showing the locations of the ROIs.
    \textbf{b--d.} Slice of the concentration field simulated
    with diffusion coefficients $D_1$, $D_2$ and $D_3$ for eight time instants covering the
    time range from 45 seconds to 790 seconds, corresponding to 100 to 1750 cardiac cycles.
    Slices are taken in the $xz$-plane ($y=0.140$ mm, the center of the geometry).
    The colorbar is scaled so the maximum is the maximum concentration of the Dendra2 
    concentration field slice at the last time instant.
    \textbf{e--g.} Mean concentration $\cbar$ in each region of interest (ROI)
    as function of time for diffusion coefficient $D_1$, $D_2$ and $D_3$, respectively. 
    \textbf{h.} The time $\hat{t}$ before the mean concentration $\cbar$ exceeds a
    threshold value of $0.25$ (ROIs 2-4) or $0.10$ (ROIs 5, 6) in each simulation setup.
    Note that there is no bar for $D_1$ in ROI 6, because the threshold value was never exceeded.}
    \label{fig:fig2}
\end{figure}

To quantify the transport of solute, we plot the profile of the mean
concentrations $\cbar_i(t)$ as functions of time
(\Cref{fig:fig2}e--g) and
the times $\hat{t}$ when the mean concentrations $\cbar_i$ first exceed
a threshold value $\hat{c}$ (\Cref{fig:fig2}h)
in 6 ROIs (\Cref{fig:fig2}a).
The dynamics at the site of photoconversion (ROI 1)
is logarithmic and follows from imposing \eqref{eq:photoconversion_curve}.
For ROIs 2--6, we observe faster kinetics with higher diffusion coefficients
(\Cref{fig:fig2}e--g),
which results in smaller times-to-threshold
(\Cref{fig:fig2}h).
For all three diffusion coefficients, the dynamics of ROI 5
is significantly delayed compared to ROI 2--4,
and ROI 6 is the slowest. Notably, for $D_1$ the mean concentration
in ROI 6 is close to zero at the end of simulation.
To quantify the balance of advection versus diffusion as transport mechanisms,
we calculate the Péclet numbers $\mathrm{Pe}_1=519$,
$\mathrm{Pe}_2=14.7$ and $\mathrm{Pe}_3=7.4$ for $D_1$, $D_2$ and $D_3$, respectively.
Since $\mathrm{Pe}>1$, this indicates that advection is the
dominant transport mechanism on a global scale.

@@

We obtained very similar kinetics curves (\Cref{fig:fig3}a, b)
and time-to-threshold values (\Cref{fig:fig3}c, d) when comparing our
experiments (\Cref{fig:fig3}a, c) with
simulation (\Cref{fig:fig3}b, d).
Importantly, since the initial condition of the simulation is a mean concentration of zero,
the mean concentration $\cbar$ can be interpreted as the change in mean concentration and
is thus a similar metric to $\Delta F$.
The final values of $\cbar$ in our simulations 0.76, 0.72, 0.82, 0.56 and 0.35
(ROIs 2--6, respectively)
are generally higher than the mean experimental values 0.63, 0.71, 0.67, 0.29 and 0.18 for $\Delta F$.
Notwithstanding similar dynamics, we observe small differences between the experiments and simulation.
For instance, transport in ROIs 2--6 is faster in the simulation and reaches higher final values.
Moreover, the $\cbar$ profiles of ROIs 3--4 are opposite order in
the simulations as compared to the experiments.
Despite observing small differences, the \emph{in vivo} experiments and
simulations are similar enough to convincingly validate the model.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure3_compare_experiments_and_simulations.png}
    \caption{
    \textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$ 
    in each region of interest (ROI)
    observed during experiments of photoconversion of Dendra2 proteins in ROI 1.
    \textbf{b.} Mean concentration $\cbar(t)$ in each ROI in a
    simulation of Dendra2 protein transport with
    the diffusion coefficient $D_3$.
    \textbf{c.} Time to $\Delta F(t)$ reaches a threshold value of 0.25 in
    one representative zebrafish larva.
    Time to $\cbar(t)$ reaches a threshold value of 0.25 for
    a transport simulation with diffusion coefficient $D_3$.
    We have excluded the posterior-most regions of the geometries,
    because the values of $\Delta F$ and $\cbar$ never exceeded the threshold
    value of 0.25 in these regions.}
    \label{fig:fig3}
\end{figure}
% Final Delta F values
%0.6322650156762044 ROI 2
%0.7059449124175479 ROI 3
%0.6694500367160006 ROI 4
%0.28986700094045437 ROI 5
%0.18155055624229066 ROI 6
% Final mean c values ROIs 2-6: 0.7637806  0.72025859 0.81903216 0.56204164 0.34926411

\subsection*{Absence of ciliary motion affects local protein distribution}
Since CSF flow is highly disturbed in the absence of motile cilia, we
want to assess the impact of cilia motion on solute distribution. To this end,
we measured the movement of photoconverted Dendra2 signal
in control and motile-cilia mutant (\emph{smh}) zebrafish .
By plotting the means of $\Delta F$ (\Cref{fig:fig4}a)
and the time to threshold
across the control and mutant cohorts (\Cref{fig:fig4}b),
we identified a significant
difference in Dendra2 distribution in ROIs 2 and 4, but not in ROIs 3, 5, and 6. In ROIs 2 and 4,
which measure Dendra2 distribution downstream of the photoconversion site,
we observed a slower increase of fluorescence intensity,
resulting in $45\%$ (ROI 2) and $25\%$ (ROI 4)
higher mean times-to-threshold in the \emph{smh}
mutants compared to controls (\Cref{fig:fig4}b).

Next, to identify whether our model would recapitulate these findings,
we simulate protein transport by using the only-cardiac/no-cilia flow model 
and compare it with the baseline model
(both cilia and cardiac motion) (\Cref{fig:fig4}c).
Similar to our experiments, we observe slower dynamics in simulations for ROIs 2 and 4,
but small changes for ROIs 3 and 6.
Interestingly, the difference was even greater for the simulations,
with the times-to-threshold in ROI 2 and 4 being 90\% and 34\% higher
in the absence of cilia-mediated flow.
Moreover, ROI 5 exhibits very different $\cbar$ profiles for the two
configurations in our computational model,
but not in our experimental data.
Altogether, our findings identify that diffusion plays an important
role for solute movement to regions upstream in the cilia flow fields or to distant ventricles.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $F(t)$
    in each region of interest (ROI) for control (black lines)
    and \emph{smh} mutant (red lines) fish. 
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for control (gray dots) and
    \emph{smh} mutant (red dots) fish.
    The thick black lines denote mean values of the datasets. The $p$-values
    were calculated using a Mann-Whitney U test with a confidence level of 95\%, with 
    the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-only/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in both panels \textbf{a} and \textbf{c}, the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

\subsection*{Loss of cilia motility in specific subpopulation slows down transport}
We previously identified three distinct ciliated cell lineages in the
zebrafish larval brain~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain}.
To identify the role of these lineages on CSF transport, we simulate transport
upon region-specific cilia paralysis by modifying the boundary conditions
of the CSF flow model. Three cilia-covered regions were considered separately:
the telencephalic (anterior) ventricle, and the dorsal and ventral parts
of the diencephalic (middle) ventricle (\Cref{fig:fig5}a).
The cilia tangential traction boundary condition is for each case
changed to a free-slip boundary condition in the paralyzed region,
prescribing zero tangential forces to
represent absence of cilia motion.
We simulate CSF flow modeled by the Stokes equations with the
modified boundary conditions and use the resulting
velocity fields in transport simulations.

The CSF velocities are most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity magnitude drops from 27.7 \textmu m/s to 8.0 \textmu m/s.
\emph{In vivo}, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic cilia has minimal
impact and results in maximum speeds of 27.7 \textmu m/s or 27.2 \textmu m/s, respectively.
Changes in the transport dynamics of Dendra2 are most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit changes in the dynamics of $\cbar$ (\Cref{fig:fig5}b).
Compared to the baseline model results, we observe a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
In contrast, removal of telencephalic cilia does not alter the evolution of the
mean concentration $\cbar$ with time (\Cref{fig:fig5}b).
Finally, without ventral cilia in the diencephalic ventricle,
more solute spreads towards the anterior ventricle (ROI 5) and
less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions where loss of cilia motility is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    loss-of-cilia-motility scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

\subsection*{Modification of ventricles geometry impacts protein distribution}
Variations in ventricular morphology impact CSF flow patterns.
To assess how ventricular morphologies influence solute distribution,
we consider four alternative brain ventricles meshes:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Cref{fig:fig6}a).
We simulate CSF flow in the modified geometries with the baseline flow model
and use the resulting velocity field to simulate transport of Dendra2
(\Cref{fig:fig6}b).

Constricting the connection between the anterior and middle ventricle leads to
slower transport to the anterior ventricle (ROI 5),
and a slightly higher mean concentration in ROIs 2--4 and 6, compared to the original geometry.
On the other hand, constricting the duct connecting the middle and the posterior ventricle
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a shrunk middle ventricle.
Additionally, for the latter configuration, the transport upstream of CSF flow towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.
For the mesh where all three ventricles are modified,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while both the anterior and the posterior ventricles (ROIs 5 and 6) have lower mean concentrations.
Altogether, our results show that the geometry and width of the ventricular ducts matter
for solute transport. This may explain why we observed larger differences
in the cilia-deficient computational model compared to the experimental data in
\Cref{fig:fig4}, as the model and the fish have different duct diameters.
